\section{Experimental Setup}
\label{sec:setup}

\subsection{MSI Dataset}
\paragraph{Mass spectrometry imaging with ion-mobility}
Mass spectrometry measures ions that are derived from the molecules present in a complex biological sample.
These spectra can be acquired at each location (pixel) of a heterogeneous sample, allowing for collection of spatially
resolved mass spectra.
This mode of analysis is known as \textit{mass spectrometry imaging (MSI)}.
The addition of \textit{ion-mobility separation (IMS)} to MSI adds another dimension, drift time
The combination of IMS with MSI is finding increasing applications in the study of disease diagnostics, plant
engineering, and microbial interactions. Unfortunately, the scale of MSI data and complexity of analysis presents a significant challenge to
scientists: a single 2D-image may be many gigabytes and comparison of multiple images is beyond the capabilities
available to many scientists. The addition of IMS exacerbates these problems.

\paragraph{Utility of CX/PCA in MSI}
Dimensionality reduction techniques can help reduce MSI datasets to more amenable sizes.
Typical approaches for dimensionality reduction include PCA and NMF, but interpretation of the results is difficult
because the components extracted via these methods are typically combinations of many hundreds or thousands of features in the original data.
CX decompositions circumvent this problem by identifying small numbers of columns in the
original data that reliably explain a large portion of the variation in the data.
This facilitates rapidly pinpointing important ions and locations in MSI applications.

In this paper, we analyze one of the largest (1TB sized) mass-spec
imaging datasets in the field. The sheer size %and complexity 
of this dataset
has previously made complex analytics intractable. This paper presents first-time science results from the successful application of CX to TB-sized data.

\subsection{Platforms}

  \begin{table*}
    \begin{center}
    \begin{tabular}{| l | c | c | c | c | c | c | c |}
    \toprule
    \textbf{Platform} & \textbf{Total Cores} & \textbf{Core Frequency} & \textbf{Interconnect} & \textbf{DRAM} & \textbf{SSDs} \\
    \midrule
    Amazon EC2 \texttt{r3.8xlarge} & 960 (32 per-node) & 2.5 GHz & 10 Gigabit Ethernet & 244 GiB & 2 x 320 GB \\
    \midrule
    Cray XC40 & 960 (32 per-node) & 2.3 GHz & Cray Aries~\cite{alverson2012cray,craycascadesc12} & 252 GiB & None \\
    \midrule
    Experimental Cray cluster & 960 (24 per-node) & 2.5 GHz & Cray Aries~\cite{alverson2012cray,craycascadesc12} & 126 GiB & 1 x 800 GB \\
    \bottomrule
    \end{tabular}
    \end{center}
    \caption{Specifications of the three hardware platforms used in these performance experiments.}
    \label{tab:hwspecs}
  \end{table*}

 In order to assess the relative performance of CX matrix factorization on various hardware, we choose the following contemporary platforms:
 \begin{itemize}
 \item a Cray\textregistered~XC40\textsuperscript{\tiny\texttrademark}
 system~\cite{alverson2012cray,craycascadesc12},
 \item an experimental Cray cluster, and
 \item an Amazon EC2 \texttt{r3.8xlarge} cluster.
 \end{itemize}

 For all platforms, we sized the Spark job to use 960 executor cores (except as
 otherwise noted).  Table~\ref{tab:hwspecs} shows the full specifications of
 the three platforms. Note that these are state-of-the-art configurations in
 datacenters and high performance computing centers. 

